%%==================================================
%% abstract.tex for XMU Graduate Thesis
%%==================================================

\begin{abstract}

近年来，人工智能，特别是大模型技术的发展，为材料科学中的知识获取、规律发现和辅助分析提供了新的技术手段。碳纳米管阵列因具备优良的电学、热学和力学性能，在储能器件、热界面材料、场发射器件及高性能复合材料等领域具有广泛的应用前景。然而，碳纳米管阵列研究具有较强的专业性，相关机理也较为复杂。对于这类科研场景，通用大模型往往受限于领域知识覆盖不足以及事实约束能力有限，容易产生知识幻觉，导致证据支撑不足、分析结果稳定性较差，因而难以满足科研任务对准确性、可追溯性和可解释性的要求。与此同时，碳纳米管阵列的制备过程受温度、气体流量、催化剂状态、生长时间等多种工艺参数共同影响，其形貌演化规律较为复杂；扫描电子显微镜图像的分析又在较大程度上依赖人工经验，工艺参数与形貌特征之间的映射关系也难以准确描述。基于此，本文围绕碳纳米管阵列研究中的知识增强、关键形貌表征与关系建模开展研究，主要内容如下。

（1）针对碳纳米管领域知识来源分散、异构性强、难以直接支撑科研分析的问题，提出一种面向碳纳米管阵列研究的多源知识建模与检索增强方法。面向学术文献、实验记录、工艺参数表、结构化实验数据以及图像分析结果，构建统一的知识规范化处理流程；提出基于层次化切分的最小语义事实单元表示方法，建立工艺参数、形貌特征、生长机理与文献证据之间的关联结构；在此基础上，进一步设计面向科研任务的关系约束检索算法，实现工艺分析、形貌解释和预测解释等任务中的证据链检索与结果追溯。

（2）针对碳纳米管阵列扫描电子显微镜图像形貌复杂、人工判读成本较高且一致性不足的问题，研究自动化关键形貌表征方法。基于前景分割结果，结合骨架分析与几何计算，构建样品级量化流程，提取密度、取向度、表观直径、平均曲率、波曲度和迂曲度等关键特征；将原本依赖经验判断的形貌信息转化为可比较的结构化表征，并通过与人工分级结果的一致性比较验证该方法的合理性和有效性。

（3）针对碳纳米管阵列制备过程中工艺参数与关键形貌之间关系复杂、结果解释能力不足的问题，建立知识增强支持下的工艺—形貌关系分析与预测方法。在知识检索结果的支持下，对工艺变量的分析范围进行收缩，重点围绕 Fe 功率、Fe 厚度和退火时长开展统计关系分析与多特征建模；通过构建工艺参数与取向度、平均曲率、波曲度和迂曲度之间的非线性映射关系，实现对关键形貌指标的预测，并结合知识库中的机理证据对变量作用及预测结果进行辅助解释，从而提升分析结果的可理解性与可追溯性。

（4）围绕碳纳米管阵列研究中的集成化应用需求，设计并实现科研辅助分析平台。平台集成多源数据管理、知识检索增强、图像处理和预测分析等核心模块，形成从知识组织、关键形貌表征到关系分析与结果解释的一体化研究流程。实验结果表明，本文方法能够实现多源知识的统一组织、关键形貌的自动化量化以及工艺参数驱动的形貌关系分析与预测，并可为分析工艺参数变化与阵列形貌演化之间的关系提供支持。

\end{abstract}

\keywords{碳纳米管阵列；知识增强；形貌表征；机器学习}

\begin{enabstract}

In recent years, the development of artificial intelligence, especially large model technology, has provided new technical means for knowledge acquisition, pattern discovery, and assisted analysis in materials science. Carbon nanotube (CNT) arrays, with their excellent electrical, thermal, and mechanical properties, have broad application prospects in energy storage devices, thermal interface materials, field emission devices, and high-performance composites. However, research on CNT arrays is highly specialized and mechanism-complex. In such scientific scenarios, general large models are often limited by insufficient domain knowledge coverage and weak factual constraints, which may lead to knowledge hallucination, inadequate evidence support, and poor stability of analytical results. As a result, they cannot readily satisfy the requirements of scientific research for accuracy, traceability, and interpretability. Meanwhile, the preparation of CNT arrays is jointly affected by multiple process parameters, such as temperature, gas flow rate, catalyst state, and growth time, making morphology evolution highly complex. In addition, the analysis of scanning electron microscopy (SEM) images still depends heavily on manual experience, and the mapping relationship between process parameters and morphology features remains difficult to characterize accurately. Therefore, this thesis focuses on knowledge enhancement, key morphology characterization, and relation modeling in CNT array research. The main contributions are as follows.

(1) To address the problem that domain knowledge in CNT research is scattered, heterogeneous, and difficult to directly support scientific analysis, a multi-source knowledge modeling and retrieval-augmented method for CNT array research is proposed. A unified normalization process is constructed for academic literature, experimental records, process parameter tables, structured experimental data, and image analysis results. A Minimum Semantic Fact Unit representation based on hierarchical segmentation is proposed, and an association structure among process parameters, morphology features, growth mechanisms, and literature evidence is established. On this basis, a relation-constrained retrieval algorithm for scientific research tasks is further designed to support evidence-chain retrieval and result traceability in process analysis, morphology interpretation, and prediction explanation tasks.

(2) To address the problem that SEM images of CNT arrays exhibit complex morphologies and that manual interpretation is costly and limited in consistency, an automated key morphology characterization method is studied. Based on foreground segmentation results and combined with skeleton analysis and geometric computation, a sample-level quantification pipeline is constructed to extract key features including density, alignment, apparent diameter, average curvature, waviness, and tortuosity. This method transforms morphology information that originally relied on empirical judgment into comparable structured representations, and its rationality and effectiveness are validated through consistency comparison with manual grading results.

(3) To address the problem that the relationships between process parameters and key morphologies in CNT array preparation are complex and that the interpretability of analytical results is insufficient, a knowledge-enhanced process-morphology relation analysis and prediction method is established. With the support of retrieval results, the scope of process-variable analysis is narrowed, focusing on Fe power, Fe thickness, and annealing time for statistical relationship analysis and multi-feature modeling. By constructing nonlinear mappings between process parameters and key morphology indicators, including alignment, average curvature, waviness, and tortuosity, the prediction of key morphology features is achieved. In addition, mechanism evidence retrieved from the knowledge base is incorporated to explain variable effects and prediction results, thereby improving the interpretability and traceability of the analysis.

(4) To meet the integrated application requirements in CNT array research, a scientific research assistant analysis platform is designed and implemented. The platform integrates core modules including multi-source data management, knowledge retrieval augmentation, image processing, and predictive analysis, forming an integrated research workflow from knowledge organization and key morphology characterization to relation analysis and result interpretation. Experimental results show that the proposed methods can achieve unified organization of multi-source knowledge, automated quantification of key morphologies, and process-parameter-driven process-morphology relation analysis and prediction, thereby providing support for analyzing the relationship between process-parameter variation and morphology evolution in CNT arrays.

\end{enabstract}

\enkeywords{Carbon Nanotube Arrays; Knowledge Enhancement; Relation Modeling; Key Morphology Characterization; Machine Learning}
